---
title: "Surveillance curves"
output:
    html_document:
      code_folding: show
      theme: united
      toc: true
      toc_depth: 3
      toc_float: true
      number_section: true

---

```{css, echo=FALSE}
.header-section-number::after {
  content: ".";
}
```

Packages used in work.
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)

```

```{r echo = FALSE, message=FALSE, warning=FALSE}
# Absent packages will be installed
if (!require(ggplot2)) install.packages("ggplot2")
if (!require(surveillance)) install.packages("surveillance")
if (!require(survival)) install.packages("survival")
if (!require(dplyr)) install.packages("dplyr")
if (!require(survminer)) install.packages("survminer")
if (!require(ranger)) install.packages("ranger")
if (!require(ggfortify)) install.packages("ggfortify")
if (!require(coin)) install.packages("coin")
```

```{r echo = T, warning=FALSE, message=FALSE, eval=TRUE}
library(surveillance)
library(survival)
library(ggplot2)
library(dplyr)
library(survminer)
library(ranger)
library(ggfortify)
library(coin)
```

```{r echo = FALSE, warning=FALSE, message=FALSE, eval=TRUE}
theme_set(theme_bw())
```

# Data desctiption
## Investigation of the dataframe

```{r}
data("ovarian")
str(ovarian)
```

The following variables are presented in this dataset:

* futime:	survival or censoring time
* fustat:	censoring status
* age:	in years
* resid.ds:	residual disease present (1=no,2=yes)
* rx:	treatment group
* ecog.ps:	ECOG performance status (1 is better, see reference)

Transform variables `resid.ds`, `rx`, `ecog.ps` into factors.
```{r}
ovarian <- ovarian %>% mutate(resid.ds = as.factor(resid.ds),
                   rx = as.factor(rx),
                   ecog.ps = as.factor(ecog.ps))
```

Rename factor's level.
```{r}
levels(ovarian$resid.ds) <- c("No", "Yes")
```

# EDA
## Number of censoring values
```{r}
ovarian %>% group_by(fustat) %>% dplyr::summarise(n = n())
```

There are more than a half censoring values which is not favorable for further analysis.

## Number of observations of different treatment type
```{r}
ovarian %>% group_by(rx) %>% dplyr::summarise(n = n())
```

There are equal group number.

## Age range
```{r}
ovarian %>% dplyr::summarise(min_age = round(min(ovarian$age), 2),
                             max_age = round(max(ovarian$age), 2),
                             mean_age = round(mean(ovarian$age), 2))
```

## Distribution of patient's ages and censoring status
It could be seen that many patients, whose age was above 60, left the experiment.
```{r}
ggplot(ovarian, aes(x = factor(fustat), y = age, fill = factor(fustat)))+
  geom_boxplot(width = 0.4) + 
  geom_jitter(size = 0.8) +
  labs(title = "Distribution of patient's ages and censoring status") +
  scale_fill_brewer(palette = "Set2") +
  guides(fill = guide_legend(title = "Censoring status")) +
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5))
```

## Distribution of patient's ages and treatment type
```{r}
ggplot(ovarian, aes(x = rx, y = age, fill = rx))+
  geom_boxplot(width = 0.4) + 
  geom_jitter(size = 0.8) +
  labs(title = "Distribution of patient's ages and treatment type") +
  scale_fill_brewer(palette = "Set2") + 
  guides(fill = guide_legend(title = "Treatment")) + 
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5))
```

## Surveillance for patients of different ages and treatment
```{r warning=FALSE, message=FALSE}
ggplot(ovarian, aes(x = age, y = futime, color = rx, fill = rx))+
  geom_point() +
  geom_smooth(alpha = 0.2) +
  labs(title = "Surveillance for patients of different ages depending on the treatment type") +
  scale_color_brewer(palette = "Set2") + 
  scale_fill_brewer(palette = "Set2") + 
  guides(fill = guide_legend(title = "Treatment"), color = guide_legend(title = "Treatment")) +
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5))
```

## Presence of the residual disease
The plot below demonstrates that there may be some relationship between age and the presence of the residual disease.
```{r}
ggplot(ovarian, aes(x = resid.ds, y = age, fill = resid.ds))+
  geom_boxplot() +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Presence of the residual disease depending on patients' age") +
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5))
```

# Kaplan–Meier curves
```{r}
km <- with(ovarian, Surv(futime, fustat))
km
```

```{r}
km_fit_1 <- survfit(Surv(futime, fustat) ~ 1, data = ovarian)
autoplot(km_fit_1)
```

Let's make additional variables for further plotting.
```{r}
ovarian <-ovarian %>% mutate(age_group = ifelse((age < mean(ovarian$age)), "Below_average", "Above_average"),
                             age_group = factor(age_group))

ovarian$age_treatment <- factor(apply(ovarian[, c(7,5)], 1, paste, collapse = "_"))

ovarian$age_resid <- factor(apply(ovarian[, c(7,4)], 1, paste, collapse = "_"))

ovarian$resid_rx <- factor(apply(ovarian[, c(4,5)], 1, paste, collapse = "_"))
```

```{r}
# rx
km_fit_rx <- survfit(Surv(futime, fustat) ~ rx, data = ovarian)
autoplot(km_fit_rx) + 
  scale_color_brewer(palette = "Set2") + 
  scale_fill_brewer(palette = "Set2") + 
  guides(fill = guide_legend(title = "Treatment type"), color = guide_legend(title = "Treatment type")) +
  labs(title = "Surveillance curves for patients with different treatment type") +
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5))

# age group
km_fit_age <- survfit(Surv(futime, fustat) ~ age_group, data = ovarian)
autoplot(km_fit_age) +
  scale_color_brewer(palette = "Set2") + 
  scale_fill_brewer(palette = "Set2") +
  guides(fill = guide_legend(title = "Age group"), color = guide_legend(title = "Age group")) +
  labs(title = "Surveillance curves for patients of different age groups") +
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5))

# age & rx
km_fit_age_rx <- survfit(Surv(futime, fustat) ~ age_treatment, data = ovarian)
autoplot(km_fit_age_rx) +
  scale_color_brewer(palette = "Dark2") + 
  scale_fill_brewer(palette = "Dark2") +
  guides(fill = guide_legend(title = "Age and \nTreatment type"), color = guide_legend(title = "Age and \nTreatment type")) +
  labs(title = "Surveillance curves for patients of different \nage groups and treatment type") +
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5))

# age & resid.ds
km_fit_age_resid <- survfit(Surv(futime, fustat) ~ age_resid, data = ovarian)
autoplot(km_fit_age_resid) +
  scale_color_brewer(palette = "Dark2") + 
  scale_fill_brewer(palette = "Dark2") +
  guides(fill = guide_legend(title = "Age and \nResidual disease"), color = guide_legend(title = "Age and \nResidual disease")) + 
  labs(title = "Surveillance curves for patients of different \nage groups and presence of residual disease") +
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5))

# resid.ds & rx
km_fit_rx_resid <- survfit(Surv(futime, fustat) ~ resid_rx, data = ovarian)
autoplot(km_fit_rx_resid) +
  scale_color_brewer(palette = "Dark2") + 
  scale_fill_brewer(palette = "Dark2") +
  guides(fill = guide_legend(title = "Residual disease and \nTreatment type"), color = guide_legend(title = "Residual disease and \nTreatment type")) +
  labs(title = "Surveillance curves for patients with different \ntreatment type and presence of residual disease") +
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5))
```

# Log-rank tests
Log-rank tests are used to compare the survival distributions of two samples.
```{r}
survdiff(Surv(futime, fustat) ~ rx, data = ovarian) # No statistical difference between survival curves

survdiff(Surv(futime, fustat) ~ age_group, data = ovarian) # No statistical difference between survival curves

survdiff(Surv(futime, fustat) ~ resid.ds, data = ovarian) # No statistical difference between survival curves

survdiff(Surv(futime, fustat) ~ ecog.ps, data = ovarian) # No statistical difference between survival curves
```

# Cox model
Cox model is used for investigating the association between the survival time of patients and one or more predictor variables (risks).
```{r}
cox <- coxph(Surv(futime, fustat) ~ rx + age_group + resid.ds + ecog.ps, data = ovarian)
summary(cox)

cox_fit <- survfit(cox)
autoplot(cox_fit)
```

Here we can see that there is no factors affecting survival time.
