---
title: "Surveillance curves"
output:
    html_document:
      code_folding: show
      theme: united
      toc: true
      toc_depth: 3
      toc_float: true
      number_section: true

---

```{css, echo=FALSE}
.header-section-number::after {
  content: ".";
}
```

Packages used in work.
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)

```

```{r echo = FALSE, message=FALSE, warning=FALSE}
# Absent packages will be installed
if (!require(ggplot2)) install.packages("ggplot2")
if (!require(surveillance)) install.packages("surveillance")
if (!require(survival)) install.packages("survival")
if (!require(dplyr)) install.packages("dplyr")
if (!require(survminer)) install.packages("survminer")
if (!require(ranger)) install.packages("ranger")
if (!require(ggfortify)) install.packages("ggfortify")
if (!require(coin)) install.packages("coin")
```

```{r echo = T, warning=FALSE, message=FALSE, eval=TRUE}
library(surveillance)
library(survival)
library(ggplot2)
library(dplyr)
library(survminer)
library(ranger)
library(ggfortify)
library(coin)
```

```{r echo = FALSE, warning=FALSE, message=FALSE, eval=TRUE}
theme_set(theme_bw())
```

# Data desctiption
## Investigation of the dataframe

```{r}
data("ovarian")
str(ovarian)
```

The following variables are presented in this dataset:

* futime:	survival or censoring time
* fustat:	censoring status
* age:	in years
* resid.ds:	residual disease present (1=no,2=yes)
* rx:	treatment group
* ecog.ps:	ECOG performance status **

** *ECOG performance status*

**0** 	Fully active, able to carry on all pre-disease performance without restriction   
**1** 	Restricted in physically strenuous activity but ambulatory and able to carry out work of  a light or sedentary nature, e.g., light house work, office work    
**2** 	Ambulatory and capable of all selfcare but unable to carry out any work activities; up and about more than 50% of waking hours     
**3** 	Capable of only limited selfcare; confined to bed or chair more than 50% of waking hours   
**4** 	Completely disabled; cannot carry on any selfcare; totally confined to bed or chair  
**5** 	Dead

Transform variables `resid.ds`, `rx`, `ecog.ps` into factors.
```{r}
ovarian <- ovarian %>% mutate(resid.ds = as.factor(resid.ds),
                   rx = as.factor(rx),
                   ecog.ps = as.factor(ecog.ps))
```

Rename factor's level.
```{r}
levels(ovarian$resid.ds) <- c("No", "Yes")
```

# EDA
## Number of censoring values
```{r}
ovarian %>% group_by(fustat) %>% dplyr::summarise(n = n())
```

There are more than a half censoring values which is not favorable for further analysis.

## Number of observations of different treatment type
```{r}
ovarian %>% group_by(rx) %>% dplyr::summarise(n = n())
```

## Number of patients with different ECOG status
```{r}
ovarian %>% group_by(ecog.ps) %>% dplyr::summarise(n = n())
```

There are equal group number.

## Age range
```{r}
ovarian %>% dplyr::summarise(min_age = round(min(ovarian$age), 2),
                             max_age = round(max(ovarian$age), 2),
                             mean_age = round(mean(ovarian$age), 2))
```

## Distribution of patient's ages and censoring status
It could be seen that many patients, whose age was above 60, left the experiment.
```{r}
ggplot(ovarian, aes(x = factor(fustat), y = age, fill = factor(fustat)))+
  geom_boxplot(width = 0.4) + 
  geom_jitter(size = 0.8) +
  labs(title = "Distribution of patient's ages and censoring status") +
  scale_fill_brewer(palette = "Set2") +
  guides(fill = guide_legend(title = "Censoring status")) +
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5))
```

## Distribution of patient's ages and treatment type
The age range in the second treatment type group was less than in the first treatment type group.
```{r}
ggplot(ovarian, aes(x = rx, y = age, fill = rx))+
  geom_boxplot(width = 0.4) + 
  geom_jitter(size = 0.8) +
  labs(title = "Distribution of patient's ages and treatment type") +
  scale_fill_brewer(palette = "Set2") + 
  guides(fill = guide_legend(title = "Treatment")) + 
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5))
```

## Surveillance for patients of different ages and treatment
```{r warning=FALSE, message=FALSE}
ggplot(ovarian, aes(x = age, y = futime, color = rx, fill = rx))+
  geom_point() +
  geom_smooth(alpha = 0.2) +
  labs(title = "Surveillance for patients of different ages \ndepending on the treatment type") +
  scale_color_brewer(palette = "Set2") + 
  scale_fill_brewer(palette = "Set2") + 
  guides(fill = guide_legend(title = "Treatment"), color = guide_legend(title = "Treatment")) +
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5))
```

## Presence of the residual disease depending on patients' age
The plot below demonstrates that there may be some relationship between patients' age and the presence of the residual disease.
```{r}
ggplot(ovarian, aes(x = resid.ds, y = age, fill = resid.ds))+
  geom_boxplot(width = 0.4) + 
  geom_jitter(size = 0.8) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Presence of the residual disease depending on patients' age") +
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5))
```

## ECOG status depending on patients' age and type of treatment

```{r}
ggplot(ovarian, aes(x = ecog.ps, y = age, fill = ecog.ps))+
  geom_boxplot(width = 0.4) + 
  geom_jitter(size = 0.8) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "ECOG status depending on patients' age") +
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5))
```

# Kaplan–Meier curves
```{r}
km <- with(ovarian, Surv(futime, fustat))
km
```
## General curve
```{r}
km_fit_1 <- survfit(Surv(futime, fustat) ~ 1, data = ovarian)
autoplot(km_fit_1) +
  labs(title = "General surveillance curve") +
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5))
```
## Additional variables
Let's make additional variables for further plotting.
```{r}
ovarian <-ovarian %>% mutate(age_group = ifelse((age < mean(ovarian$age)), "Below_average", "Above_average"),
                             age_group = factor(age_group))

ovarian$age_treatment <- factor(apply(ovarian[, c(7,5)], 1, paste, collapse = "_"))

ovarian$age_resid <- factor(apply(ovarian[, c(7,4)], 1, paste, collapse = "_"))

ovarian$resid_rx <- factor(apply(ovarian[, c(4,5)], 1, paste, collapse = "_"))

ovarian$ecog_treatment <- factor(apply(ovarian[, c(6,5)], 1, paste, collapse = "_"))

ovarian$ecog_age <- factor(apply(ovarian[, c(7,6)], 1, paste, collapse = "_"))

ovarian$ecog_resid <- factor(apply(ovarian[, c(6,4)], 1, paste, collapse = "_"))
```
## Treatment type
On the plot below we can notice that the surveillance of patients from the second treatment type group has a tendency to be higher comparing to the first treatment type group.
```{r}
# rx
km_fit_rx <- survfit(Surv(futime, fustat) ~ rx, data = ovarian)
autoplot(km_fit_rx) + 
  scale_color_brewer(palette = "Set2") + 
  scale_fill_brewer(palette = "Set2") + 
  guides(fill = guide_legend(title = "Treatment type"), color = guide_legend(title = "Treatment type")) +
  labs(title = "Surveillance curves for patients from \ndifferent treatment type groups") +
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5))
```

## ECOG status
Here we can see the declining tendency for patients with *2* ECOG status to survive. 
```{r}
# ecog.ps
km_fit_rx <- survfit(Surv(futime, fustat) ~ ecog.ps, data = ovarian)
autoplot(km_fit_rx) + 
  scale_color_brewer(palette = "Set2") + 
  scale_fill_brewer(palette = "Set2") + 
  guides(fill = guide_legend(title = "ECOG status"), color = guide_legend(title = "ECOG status")) +
  labs(title = "Surveillance curves for patients with different ECOG status") +
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5))
```

## Age group
The plot below shows that there is a tendency of survival rates to be greater for patients whose age is below the average.
```{r}
# age group
km_fit_age <- survfit(Surv(futime, fustat) ~ age_group, data = ovarian)
autoplot(km_fit_age) +
  scale_color_brewer(palette = "Set2") + 
  scale_fill_brewer(palette = "Set2") +
  guides(fill = guide_legend(title = "Age group"), color = guide_legend(title = "Age group")) +
  labs(title = "Surveillance curves for patients of different age groups") +
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5))
```

## Age and Treatment type groups
The plot presented below shows that the highest survival rate seems to be among patients whose age is below average and who receive second type of treatment, while the lowest - among patients whose age is above average and who receive first type of treatment.
```{r}
# age & rx
km_fit_age_rx <- survfit(Surv(futime, fustat) ~ age_treatment, data = ovarian)
autoplot(km_fit_age_rx) +
  scale_color_brewer(palette = "Dark2") + 
  scale_fill_brewer(palette = "Dark2") +
  guides(fill = guide_legend(title = "Age and \nTreatment type"), color = guide_legend(title = "Age and \nTreatment type")) +
  labs(title = "Surveillance curves for patients of different \nage groups and treatment type") +
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5))
```

## Age group and Residual disease
The following plot demonstrates that the highest survival rate seems to be among patients whose age is below average and who do not have residual disease. In contrast, there is a declining tendency for patients whose age is above average and who have residual disease to survive.
```{r}
# age & resid.ds
km_fit_age_resid <- survfit(Surv(futime, fustat) ~ age_resid, data = ovarian)
autoplot(km_fit_age_resid) +
  scale_color_brewer(palette = "Dark2") + 
  scale_fill_brewer(palette = "Dark2") +
  guides(fill = guide_legend(title = "Age and \nResidual disease"), color = guide_legend(title = "Age and \nResidual disease")) + 
  labs(title = "Surveillance curves for patients of different \nage groups and presence of residual disease") +
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5))
```

## Residual disease and Treatment type
Next plot shows a trend for patients who receive second type of treatment and have residual disease to die.
```{r}
# resid.ds & rx
km_fit_rx_resid <- survfit(Surv(futime, fustat) ~ resid_rx, data = ovarian)
autoplot(km_fit_rx_resid) +
  scale_color_brewer(palette = "Dark2") + 
  scale_fill_brewer(palette = "Dark2") +
  guides(fill = guide_legend(title = "Residual disease and \nTreatment type"), color = guide_legend(title = "Residual disease and \nTreatment type")) +
  labs(title = "Surveillance curves for patients with different \ntreatment type and presence of residual disease") +
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5))
```

## ECOG status and Age group
The plot presented below shows that the lowest survival rate seems to be among patients whose age is below average and whose ECOG status is "2".
```{r}
# ecog.ps & age
km_fit_ecog_age <- survfit(Surv(futime, fustat) ~ ecog_age, data = ovarian)
autoplot(km_fit_ecog_age) +
  scale_color_brewer(palette = "Dark2") + 
  scale_fill_brewer(palette = "Dark2") +
  guides(fill = guide_legend(title = "Age and \nECOG status"), color = guide_legend(title = "Age and \nECOG status")) +
  labs(title = "Surveillance curves for patients of different \nage groups and ECOG status") +
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5))
```

## ECOG status and Treatment type
Here we can see a good trend to survive for patients who receive second type of treatment and have ECOG status "1".
```{r}
# ecog.ps & rx
km_fit_ecog_rx <- survfit(Surv(futime, fustat) ~ ecog_treatment, data = ovarian)
autoplot(km_fit_ecog_rx) +
  scale_color_brewer(palette = "Dark2") + 
  scale_fill_brewer(palette = "Dark2") +
  guides(fill = guide_legend(title = "ECOG_Treatment"), color = guide_legend(title = "ECOG_Treatment")) +
  labs(title = "Surveillance curves for patients with \ndifferent ECOG status and treatment type") +
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5))
```

## ECOG status and Residual disease
The greatest survival trend is observed for patients without residual disease and ECOG status "1".
```{r}
# ecog.ps & resid.ds
km_fit_ecog_resid <- survfit(Surv(futime, fustat) ~ ecog_resid, data = ovarian)
autoplot(km_fit_ecog_resid) +
  scale_color_brewer(palette = "Dark2") + 
  scale_fill_brewer(palette = "Dark2") +
  guides(fill = guide_legend(title = "ECOG and \nResidual disease"), color = guide_legend(title = "ECOG and \nResidual disease")) +
  labs(title = "Surveillance curves for patients with different \nECOG status and presence of residual disease") +
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5))
```

# Log-rank tests
Log-rank tests are used to compare the survival distributions.   
`surfdiff` function was used when there were two levels of factor, while `pairwise_survdiff` function was used when there were more than two levels of factor, because P value adjustment methoda are provided by this function.  
No statistical difference between survival curves was observed.
```{r}
survdiff(Surv(futime, fustat) ~ rx, data = ovarian) # No statistical difference between survival curves

survdiff(Surv(futime, fustat) ~ age_group, data = ovarian) # No statistical difference between survival curves

survdiff(Surv(futime, fustat) ~ resid.ds, data = ovarian) # No statistical difference between survival curves

survdiff(Surv(futime, fustat) ~ ecog.ps, data = ovarian) # No statistical difference between survival curves

pairwise_survdiff(Surv(futime, fustat) ~ age_treatment, data = ovarian) # No statistical difference between survival curves

pairwise_survdiff(Surv(futime, fustat) ~ age_resid, data = ovarian) # No statistical difference between survival curves

pairwise_survdiff(Surv(futime, fustat) ~ resid_rx, data = ovarian) # No statistical difference between survival curves

pairwise_survdiff(Surv(futime, fustat) ~ ecog_treatment, data = ovarian) # No statistical difference between survival curves

pairwise_survdiff(Surv(futime, fustat) ~ ecog_age, data = ovarian) # No statistical difference between survival curves

pairwise_survdiff(Surv(futime, fustat) ~ ecog_resid, data = ovarian) # No statistical difference between survival curves
```


# Cox model
Cox model is used for investigating the association between the survival time of patients and one or more predictor variables (risks).
```{r}
cox <- coxph(Surv(futime, fustat) ~ rx + age_group + resid.ds + ecog.ps, data = ovarian)
summary(cox)

cox_fit <- survfit(cox)
autoplot(cox_fit) + 
  labs(title = "Cox model curve") +
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5))
  
```

```{r}
ggforest(cox, data = ovarian)
```

Here we can see that there is no factors (among `age_group`, `rx`, `ecog.ps`, `resid.ds`)  affecting survival time.
